---
title: "Reference Categories and Baseline Groups"
---

## Categorical Predictor Variables

We already talked about how to interpret coefficients in a regression model. What we have not discussed is that the interpretation that a coefficient as the expected change in the response variable for a one-unit increase in the predictor variable, holding all other predictors constant, only applies to continuous predictor variables. For categorical predictors and other kind of bounded variables, the interpretation is different.

Let's imagine we investigate the relationship between gender and political donations. We have a binary variable `gender` that takes the value of 0 for men and 1 for women. We run a regression model with `gender` as the predictor variable and `donationAmount` as the response variable:

```{r}
#| label: simulate-gender-donation
#| echo: false

set.seed(42)

n <- 500

gender <- rbinom(n, size = 1, prob = 0.5)

# Donations depend on gender: women donate more on average
donationAmount <- 100 + 50 * gender + rnorm(n, mean = 0, sd = 30)

data <- data.frame(
  gender = gender,
  donationAmount = donationAmount
)
```

```{r}
summary(lm(donationAmount ~ factor(gender), data = data))
```


The result shows that the coefficient for `gender` with the value 1 is 50. However, what does that mean? Since no one can become one unit more woman, another interpretation is needed. In case of a binary variable, the interpretation is straightforward: The coefficient of the group that is in the model specification is compared to the group that is not in the model specification. In this case, the coefficient of 50 for `gender1` means that women donate on average 50 dollars more than men. The group that is not in the model specification, in this case men, is called the reference category or baseline group. The choice of the reference category is arbitrary and does not affect the overall fit of the model, but it does affect the interpretation of the coefficients.

The logic becomes a bit more complex when we have more than two categories. For instance, let's assume we have another categorical variable called `party identification` with four possible values: `Left`, `Center`, `Right`, and `Other`. To estimate the effect of a respondent's `party identification` on `donationAmount`, we need to choose at least one of the categories as the reference category. To illustrate the problem, we specify one regression model with `Other` as the reference category and another regression model with both `Right` and `Other` as the reference categories: 
 
```{r}
#| label: simulate-party-donation
#| echo: false

set.seed(123)

n <- 600

party <- sample(
  c("Left", "Center", "Right", "Other"),
  size = n,
  replace = TRUE,
  prob = c(0.3, 0.35, 0.25, 0.1)
)

left <- ifelse(party == "Left", 1, 0)
center <- ifelse(party == "Center", 1, 0)
right <- ifelse(party == "Right", 1, 0)
other <- ifelse(party == "Other", 1, 0)

donationAmount <- 100 +
  ifelse(party == "Left", 40, 0) +
  ifelse(party == "Center", 15, 0) +
  ifelse(party == "Right", 25, 0) +
  rnorm(n, mean = 0, sd = 25)

data_party <- data.frame(
  left = left,
  center = center,
  right = right,
  other = other,
  donationAmount = donationAmount
)
```

```{r}
# Leave out "Other" as the reference category
model_other_ref <- lm(donationAmount ~ left + center + right, data = data_party)

# Leave out "Right" and "Other" as reference categories
model_ro_ref <- lm(donationAmount ~ left + center, data = data_party)

fmt_coef <- function(cs, term) {
  if (!term %in% rownames(cs)) {
    return("—")
  }
  sprintf("%.2f (%.2f)", cs[term, 1], cs[term, 2])
}

cs_other <- coef(summary(model_other_ref))
cs_ro_ref <- coef(summary(model_ro_ref))

terms <- c("Left", "Center", "Right", "Other")

table_out <- data.frame(
  Term = terms,
  `Ref:Other` = c(
    fmt_coef(cs_other, "left"),
    fmt_coef(cs_other, "center"),
    fmt_coef(cs_other, "right"),
    "ref"
  ),
  `Ref:Right_Other` = c(
    fmt_coef(cs_ro_ref, "left"),
    fmt_coef(cs_ro_ref, "center"),
    "ref",
    "ref"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  table_out,
  align = "lcc",
  caption = "Party effects with different reference categories"
)
```
The first column of the tables shows the results when `Other` is the reference category. This means the coefficients for `Left`, `Center`, and `Right` are interpreted relative to `Other` category. For instance, the coefficient for `Left` with a value of 40 means that voters who identify as `Left` donate on average 40 dollars more than voters who identify as `Other`. However, the coefficient for `Left` changes considerably when we leave out the `Right` category as well. The reason is that the coefficient for `Left` in the second column has to be interpreted relative to the `Other` and `Right` category. Voters who identify as `Right` donate on average 23 dollars more than voters who identify as `Other` (see first column) and thus have the second largest coefficient in the first column. Consequently, leaving the `Right` category out, the coefficient for `Left` in the second column has now to be compared against `Other` and the - generously donating - `Right` category. As a result, the coefficient for `Left` becomes much smaller in the second column.

## Shares

One common case that is subject to the same problem of reference categories but is rarely discussed in the literature is the case of shares. For instance, researchers are often interested in the effect of the share of the population living in urban areas, the share of the agricultural or industrial work force, the share of ethnic or religious groups, or the share of the population with a certain level of education on some outcome (to name but a few examples). In all these cases, the share variable is bounded between 0 and 1 and thus has a reference category at both ends of the spectrum. For instance, if we have a variable that measures the share of the population living in urban areas, then the reference category for a value of 0 is that no one lives in urban areas (and everybody lives in rural areas), while the reference category for a value of 1 is that everyone lives in urban areas (and no one lives in rural areas). This means that the interpretation of the coefficients for share variables can be quite complex and depends on the number of categories.

A quite prominent example is the literature on the relationship between the ideological composition of government and spending patterns. A typical setup is to use total government spending or social security spending as the response variable and the share of specific parties in government as the predictor variable. Here again, the interpretation of the coefficient for the share of specific parties in government depends on the reference category. In some specifications, only the share of left-wing parties is included as a predictor variable, while in other specifications, the share of left-wing and center or Christian parties is included. The interpretation of the coefficient for the share of left-wing parties in government changes considerably depending on whether the share of other types of parties is included in the model specification or left out as reference category. Thus, the interpretation of the coefficients for share variables is the same as for categorical variables with the difference that share variables can take on a continuous range of values between 0 and 1, while categorical variables take on discrete values.